A Decision Procedure for String Constraints with String-Integer Conversion and Flat Regular Constraints

### 2021-05-17 FMCAD2021/5_optimization.tex
optimization
Z3&mysolver result on Arith

To avoid applying the quantifier elimination algorithm of {\pa}, we consider the following under approximation of $\tau(\ivarx_{\sigma(i)}, \ldots, \ivarx_{\sigma(1)})$, namely, we additionally assume that $\ivarx_{\sigma(i)} \le 10^{u}$ for some constant bound $u \in \Nat$ with $u \ge 1$. Then $\tau(\ivarx_{\sigma(i)}, \ldots, \ivarx_{\sigma(1)})$ can be rewritten as 
%\begin{equation}\label{eqn-org}
\[\tau'(\ivarx_{\sigma(i)}, \ldots, \ivarx_{\sigma(1)}) \Def \sum_{j=1}^{i-1} a_j 10^{\ivarx_{\sigma(j)}} + \sum_{k=1}^{i-1}b_k \ivarx_{\sigma(k)} \le c - b_i  \ivarx_{\sigma(i)}.\]
%\end{equation}
Let us assume $a_{i-1} > 0$.
It is easy to observe that $c - |b_i|10^u \le c - b_i \ivarx_{\sigma(i)} \le c + |b_i|10^u$.
Then we can apply Lemma~\ref{lem:exp-ineq} to the following two inequalities to eliminate $10^{\ivarx_{\sigma(i-1)}}$,
\begin{equation}\label{eqn-lb}
\sum_{j=1}^{i-1} a_j 10^{\ivarx_{\sigma(j)}} + \sum_{k=1}^{i-1}b_k \ivarx_{\sigma(k)} \le c - |b_i|10^u
\end{equation}
and
\begin{equation}\label{eqn-ub}
\sum_{j=1}^{i-1} a_j 10^{\ivarx_{\sigma(j)}} + \sum_{k=1}^{i-1}b_k \ivarx_{\sigma(k)} \le c + |b_i|10^u.
\end{equation}
Let $\alpha_1 = \ell_{10}(c - |b_i|10^u) -  \ell_{10}(a_{i-1})$ and $\alpha_2 = \ell_{10}(c + |b_i|10^u) -  \ell_{10}(a_{i-1})$. Then from Lemma~\ref{lem:exp-ineq}, 
\begin{itemize}
\item if $\ivarx_{\sigma(i-1)} \ge B$, $\ivarx_{\sigma(i-1)} \le \alpha_1 - 1$, and $\ivarx_{\sigma(i-1)} \ge \ivarx_{\sigma(i-2)} + \delta$, then inequality (\ref{eqn-lb}), thus also $\tau'(\ivarx_{\sigma(i)}, \ldots, \ivarx_{\sigma(1)})$, is evaluated to $\ltrue$,
\item if $\ivarx_{\sigma(i-1)} \ge B$, $\ivarx_{\sigma(i-1)} \ge \alpha_2 + 2$, and $\ivarx_{\sigma(i-1)} \ge \ivarx_{\sigma(i-2)} + \delta$, then inequality (\ref{eqn-ub}), thus also $\tau'(\ivarx_{\sigma(i)}, \ldots, \ivarx_{\sigma(1)})$, is evaluated to $\lfalse$.
\end{itemize}
Therefore, $\tau'(\ivarx_{\sigma(i)}, \ldots, \ivarx_{\sigma(1)})$, thus also $\tau(\ivarx_{\sigma(i)}, \ldots, \ivarx_{\sigma(1)})$, is equivalent to 
\[
\small
\begin{array}{l}
\bigvee \limits_{p=0}^{B-1} ( \ivarx_{\sigma(i-1)} = p \wedge \tau'(\ivarx_{\sigma(i)}, \ldots, \ivarx_{\sigma(1)}) [p/\ivarx_{\sigma(i-1)}])   \\
\bigvee \big(\ivarx_{\sigma(i-1)} \ge B \wedge \ivarx_{\sigma(i-1)} \le \alpha_1  - 1  \wedge \ivarx_{\sigma(i-1)} \ge \ivarx_{\sigma(i-2)} +\delta \big)\\
%
\bigvee \bigvee \limits_{p=0}^{\delta-1} 
\left(
\begin{array}{l}
\ivarx_{\sigma(i-1)} \ge B \wedge \ivarx_{\sigma(i-1)} \le \alpha_1  -1 \ \wedge \\
 \ivarx_{\sigma(i-1)} = \ivarx_{\sigma(i-2)} +p\ \wedge\\
 \tau'(\ivarx_{\sigma(i)}, \ldots, \ivarx_{\sigma(1)})[\ivarx_{\sigma(i-2)} + p /\ivarx_{\sigma(i-1)}] 
\end{array}
\right)\\
\bigvee \bigvee \limits_{p=0}^{\delta-1} 
\left(
\begin{array}{l}
\ivarx_{\sigma(i-1)} \ge B \wedge \ivarx_{\sigma(i-1)} \ge \alpha_2 + 2 \ \wedge \\
 \ivarx_{\sigma(i-1)} = \ivarx_{\sigma(i-2)} +p\ \wedge\\
 \tau'(\ivarx_{\sigma(i)}, \ldots, \ivarx_{\sigma(1)})[\ivarx_{\sigma(i-2)} + p /\ivarx_{\sigma(i-1)}] 
\end{array}
\right)\\
\bigvee \bigvee \limits_{p = \alpha_1}^{\alpha_2 + 1}
\left(
\begin{array}{l}
\ivarx_{\sigma(i-1)} \ge B \wedge \ivarx_{\sigma(i-1)} = p\ \wedge \\
 \tau'(\ivarx_{\sigma(i)}, \ldots, \ivarx_{\sigma(1)})[p /\ivarx_{\sigma(i-1)}] 
\end{array}
\right),
\end{array}
\]
where all exponential occurrences of $\ivarx_{\sigma(i-1)}$ are eliminated.

@@

>>>>

To avoid applying the quantifier elimination algorithm of {\pa}, we consider the following under approximation of $\tau(\ivarx_{\sigma(i)}, \ldots, \ivarx_{\sigma(1)})$, namely, we additionally assume that $\ivarx_{\sigma(i)} \le 10^{u}$ for some constant bound $u \in \Nat$ with $u \ge 1$. Then $\tau(\ivarx_{\sigma(i)}, \ldots, \ivarx_{\sigma(1)})$ can be rewritten as 
%\begin{equation}\label{eqn-org}
\[\tau'(\ivarx_{\sigma(i)}, \ldots, \ivarx_{\sigma(1)}) \Def \sum_{j=1}^{i-1} a_j 10^{\ivarx_{\sigma(j)}} + \sum_{k=1}^{i-1}b_k \ivarx_{\sigma(k)} \le c - b_i  \ivarx_{\sigma(i)}.\]
%\end{equation}
Let us assume $a_{i-1} > 0$.
Define $c_1, c_2$ as follows: If $b_i > 0$, then $c_1 = c- b_i 10^u$ and $c_2 = c$, otherwise, $c_1 = c$ and $c_2 = c - b_i 10^u$.
It is easy to observe that $c_1 \le c - b_i \ivarx_{\sigma(i)} \le c_2$.
Then we can apply Lemma~\ref{lem:exp-ineq} to the following two inequalities to eliminate $10^{\ivarx_{\sigma(i-1)}}$,
\begin{equation}\label{eqn-lb}
\sum_{j=1}^{i-1} a_j 10^{\ivarx_{\sigma(j)}} + \sum_{k=1}^{i-1}b_k \ivarx_{\sigma(k)} \le c_1
\end{equation}
and
\begin{equation}\label{eqn-ub}
\sum_{j=1}^{i-1} a_j 10^{\ivarx_{\sigma(j)}} + \sum_{k=1}^{i-1}b_k \ivarx_{\sigma(k)} \le c_2.
\end{equation}
Let $\alpha_1 = \ell_{10}(c_1) -  \ell_{10}(a_{i-1})$ and $\alpha_2 = \ell_{10}(c_2) -  \ell_{10}(a_{i-1})$. Then from Lemma~\ref{lem:exp-ineq}, 
\begin{itemize}
\item if $\ivarx_{\sigma(i-1)} \ge B$, $\ivarx_{\sigma(i-1)} \le \alpha_1 - 1$, and $\ivarx_{\sigma(i-1)} \ge \ivarx_{\sigma(i-2)} + \delta$, then inequality (\ref{eqn-lb}), thus also $\tau'(\ivarx_{\sigma(i)}, \ldots, \ivarx_{\sigma(1)})$, is evaluated to $\ltrue$,
\item if $\ivarx_{\sigma(i-1)} \ge B$, $\ivarx_{\sigma(i-1)} \ge \alpha_2 + 2$, and $\ivarx_{\sigma(i-1)} \ge \ivarx_{\sigma(i-2)} + \delta$, then inequality (\ref{eqn-ub}), thus also $\tau'(\ivarx_{\sigma(i)}, \ldots, \ivarx_{\sigma(1)})$, is evaluated to $\lfalse$.
\end{itemize}
Therefore, $\tau'(\ivarx_{\sigma(i)}, \ldots, \ivarx_{\sigma(1)})$, thus also $\tau(\ivarx_{\sigma(i)}, \ldots, \ivarx_{\sigma(1)})$, is equivalent to 
\[
\small
\begin{array}{l}
\bigvee \limits_{p=0}^{B-1} ( \ivarx_{\sigma(i-1)} = p \wedge \tau'(\ivarx_{\sigma(i)}, \ldots, \ivarx_{\sigma(1)}) [p/\ivarx_{\sigma(i-1)}])   \\
\bigvee \big(\ivarx_{\sigma(i-1)} \ge B \wedge \ivarx_{\sigma(i-1)} \le \alpha_1  - 1  \wedge \ivarx_{\sigma(i-1)} \ge \ivarx_{\sigma(i-2)} +\delta \big)\\
%
\bigvee \bigvee \limits_{p=0}^{\delta-1} 
\left(
\begin{array}{l}
\ivarx_{\sigma(i-1)} \ge B \wedge \ivarx_{\sigma(i-1)} \le \alpha_1  -1 \ \wedge \\
 \ivarx_{\sigma(i-1)} = \ivarx_{\sigma(i-2)} +p\ \wedge\\
 \tau'(\ivarx_{\sigma(i)}, \ldots, \ivarx_{\sigma(1)})[\ivarx_{\sigma(i-2)} + p /\ivarx_{\sigma(i-1)}] 
\end{array}
\right)\\
\bigvee \bigvee \limits_{p=0}^{\delta-1} 
\left(
\begin{array}{l}
\ivarx_{\sigma(i-1)} \ge B \wedge \ivarx_{\sigma(i-1)} \ge \alpha_2 + 2 \ \wedge \\
 \ivarx_{\sigma(i-1)} = \ivarx_{\sigma(i-2)} +p\ \wedge\\
 \tau'(\ivarx_{\sigma(i)}, \ldots, \ivarx_{\sigma(1)})[\ivarx_{\sigma(i-2)} + p /\ivarx_{\sigma(i-1)}] 
\end{array}
\right)\\
\bigvee \bigvee \limits_{p = \alpha_1}^{\alpha_2 + 1}
\left(
\begin{array}{l}
\ivarx_{\sigma(i-1)} \ge B \wedge \ivarx_{\sigma(i-1)} = p\ \wedge \\
 \tau'(\ivarx_{\sigma(i)}, \ldots, \ivarx_{\sigma(1)})[p /\ivarx_{\sigma(i-1)}] 
\end{array}
\right),
\end{array}
\]
where all exponential occurrences of $\ivarx_{\sigma(i-1)}$ are eliminated.

@@

\paragraph{Perform the elimination of exponential occurrences of variables in parallel on atomic formulas}

Note that when Lemma~\ref{lem:exp-ineq} is applied to an atomic formula $\tau$ to eliminate exponential occurrences of $\ivarx$, it computes 4 parameters, namingly $(\alpha^\tau_1,\alpha^\tau_2,B^\tau,\delta^\tau)$, and obtains an equivalent {\pa} formula according to these parameters. Since the parameters are used as boundary conditions in enumeration, we can choose the strictest boundary conditions and apply to all atomic formulas.

More specifically, we first compute $(\alpha^\tau_1,\alpha^\tau_2,B^\tau,\delta^\tau)$ 
for all atomic formulas that contain $10^\ivarx$. Let $\alpha^{\min}_1$ denote the minimal element in $\{\alpha^\tau_1\}_\tau$, let $\alpha^{\max}_2,B^{\max},\delta^{\max}$ denote the maximal element in $\{\alpha^\tau_2\}_\tau$, $\{B^\tau\}_\tau$ and $\{\delta^\tau\}_\tau$ respectively. Then, we apply $(\alpha^{\min}_1,\alpha^{\max}_2,B^{\max},\delta^{\max})$ to all atomic formulas $\tau$ in parallel and obtain $\varphi'$ equivalent to $\varphi$.

For instance, assume coefficients of $10^\ivarx$ in atomic formulas of $\varphi$ are all non-negative. Let $\varphi^+$ denote the formula obtained by replace atomic formulas where coefficients of $10^\ivarx$ are positive by $\ltrue$. The resulted formula $\varphi'$ is
\[
\small
\begin{array}{l}
\bigvee \limits_{p=0}^{B^{\max}-1} ( \ivarx_{\sigma(i-1)} = p \wedge \varphi(\ivarx_{\sigma(i)}, \ldots, \ivarx_{\sigma(1)}) [p/\ivarx_{\sigma(i-1)}])   \\
\bigvee \left(
    \begin{array}{l}
        \ivarx_{\sigma(i-1)} \ge B^{\max} \wedge \ivarx_{\sigma(i-1)} \le \alpha^{\min}_1  - 1  \wedge \\
        \ivarx_{\sigma(i-1)} \ge \ivarx_{\sigma(i-2)} +\delta^{\max} \wedge \varphi^+  
    \end{array}
    \right)\\
%
\bigvee \bigvee \limits_{p=0}^{\delta^{\max}-1} 
\left(
\begin{array}{l}
\ivarx_{\sigma(i-1)} \ge B^{\max} \wedge \ivarx_{\sigma(i-1)} \le \alpha^{\min}_1  -1 \ \wedge \\
 \ivarx_{\sigma(i-1)} = \ivarx_{\sigma(i-2)} +p\ \wedge\\
 \varphi(\ivarx_{\sigma(i)}, \ldots, \ivarx_{\sigma(1)})[\ivarx_{\sigma(i-2)} + p /\ivarx_{\sigma(i-1)}] 
\end{array}
\right)\\
\bigvee \bigvee \limits_{p=0}^{\delta^{\max}-1} 
\left(
\begin{array}{l}
\ivarx_{\sigma(i-1)} \ge B^{\max} \wedge \ivarx_{\sigma(i-1)} \ge \alpha^{\max}_2 + 2 \ \wedge \\
 \ivarx_{\sigma(i-1)} = \ivarx_{\sigma(i-2)} +p\ \wedge\\
 \varphi(\ivarx_{\sigma(i)}, \ldots, \ivarx_{\sigma(1)})[\ivarx_{\sigma(i-2)} + p /\ivarx_{\sigma(i-1)}] 
\end{array}
\right)\\
\bigvee \bigvee \limits_{p = \alpha^{\min}_1}^{\alpha^{\max}_2 + 1}
\left(
\begin{array}{l}
\ivarx_{\sigma(i-1)} \ge B^{\max} \wedge \ivarx_{\sigma(i-1)} = p\ \wedge \\
 \varphi(\ivarx_{\sigma(i)}, \ldots, \ivarx_{\sigma(1)})[p /\ivarx_{\sigma(i-1)}] 
\end{array}
\right),
\end{array}
\]

### 2021-05-17 FMCAD2021/6_impl_exp.tex
Z3&mysolver result on Arith

% Please add the following required packages to your document preamble:
% \usepackage{multirow}
\begin{table*}[]
    \label{table:arithmetic}
    \caption{Results of Z3, CVC4 and $\paexp$-Solver on arithmetic benchmark}
    \begin{tabular}{cc|cc|cc|cc}
    \multicolumn{1}{l}{} &  & \multicolumn{2}{c|}{Z3} & \multicolumn{2}{c|}{CVC4} & \multicolumn{2}{c}{MySolver} \\
    \multicolumn{1}{l}{} &  & number & avg & number & avg & number & avg \\ \hline
    \multirow{3}{*}{\begin{tabular}[c]{@{}c@{}}unbound\\ 2-3-3-4\end{tabular}} & SAT & 31 & \textless{}0.1 & 26 & 1.2 & 35 & \textless{}0.1 \\
     & UNSAT & 37 & \textless{}0.1 & 38 & \textless{}0.1 & 65 & 0.3 \\
     & TIMEOUT & 32 &  & 36 &  & 0 &  \\ \hline
    \multirow{3}{*}{\begin{tabular}[c]{@{}c@{}}unbound\\ 3-4-4-5\end{tabular}} & SAT & 27 & \textless{}0.1 & 16 & \textless{}0.1 & 34 & 0.4 \\
     & UNSAT & 22 & \textless{}0.1 & 24 & \textless{}0.1 & 66 & 2.8 \\
     & TIMEOUT & 51 &  & 60 &  & 0 &  \\ \hline
    \multirow{3}{*}{\begin{tabular}[c]{@{}c@{}}unbound\\ 3-4-4-5 \\ with equality\end{tabular}} & SAT & 4 & 19 & 0 & \textless{}0.1 & 16 & 3.7 \\
     & UNSAT & 30 & \textless{}0.1 & 31 & 0.2 & 44 & \textless{}0.1 \\
     & TIMEOUT & 66 &  & 69 &  & 28 & 19.6 \\ \hline
    \multirow{3}{*}{\begin{tabular}[c]{@{}c@{}}unbound\\ 4-5-5-6\end{tabular}} & SAT & 20 & 0.2 & 15 & 8.2 & 33 & 5.2 \\
     & UNSAT & 22 & \textless{}0.1 & 23 & 0.2 & 50 & 4.8 \\
     & TIMEOUT & 58 &  & 62 &  & 17 &  \\ \hline\hline
    \multirow{4}{*}{\begin{tabular}[c]{@{}c@{}}bound\\ 2-3-3-4\end{tabular}} & SAT & 31 & \textless{}0.1 & 14 & \textless{}0.1 & 37 & 0,6 \\
     & UNSAT & 31 & \textless{}0.1 & 30 & \textless{}0.1 & 46 & \textless{}0.1 \\
     & B-UNSAT &  &  &  &  & 17 & 9.3 \\
     & TIMEOUT & 38 &  & 56 &  & 0 &  \\ \hline
    \multirow{4}{*}{\begin{tabular}[c]{@{}c@{}}bound\\ 3-4-4-5\end{tabular}} & SAT & 31 & 0.8 & 13 & \textless{}0.1 & 36 & 3.2 \\
     & UNSAT & 20 & \textless{}0.1 & 20 & \textless{}0.1 & 42 & \textless{}0.1 \\
     & B-UNSAT &  &  &  &  & 3 & 54 \\
     & TIMEOUT & 49 &  & 67 &  & 19 &  \\ \hline
    \multirow{4}{*}{\begin{tabular}[c]{@{}c@{}}bound\\ 3-4-4-5 \\ with equality\end{tabular}} & SAT & 3 & 23 & 1 & 39 & 14 & 27.5 \\
     & UNSAT & 25 & \textless{}0.1 & 25 & \textless{}0.1 & 47 & \textless{}0.1 \\
     & B-UNSAT &  &  &  &  & 2 & 57.9 \\
     & TIMEOUT & 72 &  & 74 &  & 37 &  \\ \hline
    \multirow{4}{*}{\begin{tabular}[c]{@{}c@{}}bound\\ 4-5-5-6\end{tabular}} & SAT & 29 & 2 & 11 & \textless{}0.1 & 28 & 10 \\
     & UNSAT & 20 & \textless{}0.1 & 20 & \textless{}0.1 & 38 & \textless{}0.1 \\
     & B-UNSAT &  &  &  &  & 0 &  \\
     & TIMEOUT & 51 &  & 69 &  & 34 & 
    \end{tabular}
\end{table*}

>>>>

% Please add the following required packages to your document preamble:
% \usepackage{multirow}
% Merge bound and unbound 
\begin{table*}[]
    \label{table:arithmetic}
    \caption{Results of Z3, CVC4 and $\paexp$-Solver on arithmetic benchmark}
    \centering
    \begin{tabular}{cc|cc|cc|cc}
        \multicolumn{1}{l}{} &  & \multicolumn{2}{c|}{Z3} & \multicolumn{2}{c|}{CVC4} & \multicolumn{2}{c}{MySolver} \\
        \multicolumn{1}{l}{} &  & number & avg & number & avg & number & avg \\ \hline
        \multirow{4}{*}{2-3-3-4} & SAT & 27 & 0.6 &  &  & 31 & 0.4 \\
         & UNSAT & 37 & \textless{}0.1 &  &  & 49 & \textless{}0.1 \\
         & B-UNSAT & \multicolumn{1}{l}{} & \multicolumn{1}{l|}{} & \multicolumn{1}{l}{} & \multicolumn{1}{l|}{} & 20 & 10.4 \\
         & TIMEOUT & 36 &  &  &  & 0 &  \\ \hline
        \multirow{4}{*}{3-4-4-5} & SAT & 20 & 0.1 &  &  & 29 & 3.5 \\
         & UNSAT & 26 & \textless{}0.1 &  &  & 42 & \textless{}0.1 \\
         & B-UNSAT & \multicolumn{1}{l}{} & \multicolumn{1}{l|}{} & \multicolumn{1}{l}{} & \multicolumn{1}{l|}{} & 1 & 53.8 \\
         & TIMEOUT & 54 &  &  &  & 28 &  \\ \hline
        \multirow{4}{*}{4-5-5-6} & SAT & 22 & 1.8 &  &  & 27 & 20.0 \\
         & UNSAT & 21 & 0.4 &  &  & 36 & 0.1 \\
         & B-UNSAT & \multicolumn{1}{l}{} & \multicolumn{1}{l|}{} & \multicolumn{1}{l}{} & \multicolumn{1}{l|}{} & 0 &  \\
         & TIMEOUT & 57 &  &  &  & 37 & 
    \end{tabular}
\end{table*}
